A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2023-08-25 paper.Rmd
formatting the paper

---
title: |
  A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol
type: ARTICLE TEMPLATE
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: a, b, c
    email: emi.tanaka@anu.edu.au
  - name: Susan VanderPlas
    affil: d
    email: susan.vanderplas@unl.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
  - num: b
    address: |
      Biological Data Science Institute, Australian National University, Acton, ACT, Australia
  - num: c
    address: |
      Research School of Finance, Actuarial Studies and Statistics, Australian National University, Acton, ACT, Australia
  - num: d
    address: |
      Department of Statistics, University of Nebraska, Lincoln, Nebraska, USA
bibliography: ref.bib
abstract: |
  Regression experts consistently recommend plotting residuals for model diagnosis, despite the availability of many numerical hypothesis test procedures designed to use residuals to assess problems with a model fit. Here we provide evidence for why this is good advice using data from a visual inference experiment. We show how conventional tests are too sensitive, which means that too often the conclusion would be that the model fit is inadequate. The experiment uses the lineup protocol which puts a residual plot in the context of null plots. This helps generate reliable and consistent reading of residual plots for better model diagnosis. It can also help in an obverse situation where a conventional test would fail to detect a problem with a model due to contaminated data. The lineup protocol also detects a range of departures from good residuals simultaneously.
keywords: |
  statistical graphics; data visualization; visual inference; hypothesis testing; reression analysis; cognitive perception; simulation; practical significance; effect size
header-includes: |
  \usepackage{lscape}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
  \doublespacing
output: rticles::tf_article
editor_options: 
  chunk_output_type: console
---

@@

In addition, scatterplots of residuals plotted against the fitted values and each of the explanatory variables, can be used to scrutinize their relationships. If there are any visually discoverable patterns, the model is potentially inadequate or incorrectly specified. **Interesting features that are not directly connected to a linear model assumption can also be found from residual plots, such as the discreteness of the fitted values, outliers and influential points.** In general, one looks for noticeable departures from the model such as non-linear pattern or heteroskedasticity. A non-linear pattern would suggest that the model needs to have some additional non-linear terms. Heteroskedasticity suggests that the error is dependent on the predictors, and hence violates the independence assumption.

@@

**There are other types of residual plot to extract the available information and identify inappropriate assumptions, like the added variable plot (cite here) for studying the role of an external variable if it enters linearly into a model. Partial residual plots (cite here) have been suggested as computationally convenient substitutes for the added variable plots. For use in diagnostic procedures, several transformations of the ordinary residuals have been suggested to overcome partially some of their shortcomings. Studentized residuals (cite here) have null distributions that are independent of the scale parameters. These residuals are shown to be closely related to a mean shift model for outliers and to the residuals obtained when each case in turn is left out of the data. Alternatively. the residuals can be transformed to have a selected covariance structure: The usual suggestion is to obtain residuals with uncorrelated elements.**

>>>>

---
title: |
  A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol
type: ARTICLE TEMPLATE
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: a, b, c
    email: emi.tanaka@anu.edu.au
  - name: Susan VanderPlas
    affil: d
    email: susan.vanderplas@unl.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
  - num: b
    address: |
      Biological Data Science Institute, Australian National University, Acton, ACT, Australia
  - num: c
    address: |
      Research School of Finance, Actuarial Studies and Statistics, Australian National University, Acton, ACT, Australia
  - num: d
    address: |
      Department of Statistics, University of Nebraska, Lincoln, Nebraska, USA
bibliography: ref.bib
abstract: |
  Regression experts consistently recommend plotting residuals for model diagnosis, despite the availability of many numerical hypothesis test procedures designed to use residuals to assess problems with a model fit. Here we provide evidence for why this is good advice using data from a visual inference experiment. We show how conventional tests are too sensitive, which means that too often the conclusion would be that the model fit is inadequate. The experiment uses the lineup protocol which puts a residual plot in the context of null plots. This helps generate reliable and consistent reading of residual plots for better model diagnosis. It can also help in an obverse situation where a conventional test would fail to detect a problem with a model due to contaminated data. The lineup protocol also detects a range of departures from good residuals simultaneously.
keywords: |
  statistical graphics; data visualization; visual inference; hypothesis testing; reression analysis; cognitive perception; simulation; practical significance; effect size
header-includes: |
  \usepackage{lscape}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
  % \doublespacing
output: rticles::tf_article
editor_options: 
  chunk_output_type: console
---

@@

In addition, scatterplots of residuals plotted against the fitted values and each of the explanatory variables, can be used to scrutinize their relationships. If there are any visually discoverable patterns, the model is potentially inadequate or incorrectly specified. \textcolor{red}{Interesting features that are not directly connected to a linear model assumption can also be found from residual plots, such as the discreteness of the fitted values, outliers and influential points.} In general, one looks for noticeable departures from the model such as non-linear pattern or heteroskedasticity. A non-linear pattern would suggest that the model needs to have some additional non-linear terms. Heteroskedasticity suggests that the error is dependent on the predictors, and hence violates the independence assumption.

@@

\textcolor{red}{There are other types of residual plot to extract the available information and identify inappropriate assumptions, like the added variable plot (cite here) for studying the role of an external variable if it enters linearly into a model. Partial residual plots (cite here) have been suggested as computationally convenient substitutes for the added variable plots. For use in diagnostic procedures, several transformations of the ordinary residuals have been suggested to overcome partially some of their shortcomings. Studentized residuals (cite here) have null distributions that are independent of the scale parameters. These residuals are shown to be closely related to a mean shift model for outliers and to the residuals obtained when each case in turn is left out of the data. Alternatively. the residuals can be transformed to have a selected covariance structure: The usual suggestion is to obtain residuals with uncorrelated elements.}

### 2023-08-25 paper.tex
double space
formatting the paper

\usepackage{lscape}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\def\tightlist{}

@@

In addition, scatterplots of residuals plotted against the fitted values
and each of the explanatory variables, can be used to scrutinize their
relationships. If there are any visually discoverable patterns, the
model is potentially inadequate or incorrectly specified. In general,
one looks for noticeable departures from the model such as non-linear
pattern or heteroskedasticity. A non-linear pattern would suggest that
the model needs to have some additional non-linear terms.
Heteroskedasticity suggests that the error is dependent on the
predictors, and hence violates the independence assumption. However,
correctly judging whether NO pattern exists in a residual plot is a
difficult task for humans. Humans will almost always see a pattern (see
\citet{kahneman2011thinking}), so the question that really needs
answering is whether any pattern perceived is consistent with
randomness, purely sampling variability, or noise. It is especially
difficult to teach this to new analysts and students
\citep{loy2021bringing}. To address this, statistical tests are
available to check for non-linear patterns
\citep[e.g.][]{ramsey1969tests}, and heteroskedasticity
\citep[e.g.][]{breusch1979simple}.

Linear regression is a well-established procedure, and there is
considerable literature describing diagnostic procedures, e.g.
\citet{draper1998applied}, \citet{montgomery1982introduction},
\citet{belsley1980regression}, \citet{cook1999applied} and
\citet{cook1982residuals}. Interestingly, despite the abundance of
conventional tests, ALL of these writings advise that plotting residuals
is an essential tool for diagnosing regression model problems:

@@

\noindent The common wisdom of experts is that plotting the residuals is
indispensable for diagnosing model fits. The ubiquity of this advice is
\emph{curious}: investigating why, is the subject of this paper.

@@

>>>>

\usepackage{lscape}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\def\tightlist{}
\usepackage{setspace}
% \doublespacing

@@

In addition, scatterplots of residuals plotted against the fitted values
and each of the explanatory variables, can be used to scrutinize their
relationships. If there are any visually discoverable patterns, the
model is potentially inadequate or incorrectly specified.
\textcolor{red}{Interesting features that are not directly connected to a linear model assumption can also be found from residual plots, such as the discreteness of the fitted values, outliers and influential points.}
In general, one looks for noticeable departures from the model such as
non-linear pattern or heteroskedasticity. A non-linear pattern would
suggest that the model needs to have some additional non-linear terms.
Heteroskedasticity suggests that the error is dependent on the
predictors, and hence violates the independence assumption.

However, correctly judging whether NO pattern exists in a residual plot
is a difficult task for humans. Humans will almost always see a pattern
(see \citet{kahneman2011thinking}), so the question that really needs
answering is whether any pattern perceived is consistent with
randomness, purely sampling variability, or noise. It is especially
difficult to teach this to new analysts and students
\citep{loy2021bringing}. To address this, statistical tests are
available to check for non-linear patterns
\citep[e.g.][]{ramsey1969tests}, and heteroskedasticity
\citep[e.g.][]{breusch1979simple}.

\textcolor{red}{There are other types of residual plot to extract the available information and identify inappropriate assumptions, like the added variable plot (cite here) for studying the role of an external variable if it enters linearly into a model. Partial residual plots (cite here) have been suggested as computationally convenient substitutes for the added variable plots. For use in diagnostic procedures, several transformations of the ordinary residuals have been suggested to overcome partially some of their shortcomings. Studentized residuals (cite here) have null distributions that are independent of the scale parameters. These residuals are shown to be closely related to a mean shift model for outliers and to the residuals obtained when each case in turn is left out of the data. Alternatively. the residuals can be transformed to have a selected covariance structure: The usual suggestion is to obtain residuals with uncorrelated elements.}

There is sustainable literature describing diagnostic procedures for
linear models, e.g. \citet{draper1998applied},
\citet{montgomery1982introduction}, \citet{belsley1980regression},
\citet{cook1999applied} and \citet{cook1982residuals}. Interestingly,
despite the abundance of conventional tests, ALL of these writings
advise that plotting residuals is an essential tool for diagnosing
regression model problems:

@@

\noindent The common wisdom of experts is that plotting the residuals is
indispensable for diagnosing model fits. The lack of empirical evidence
for the ubiquitous advice is \emph{curious}. The research study
discussed in this paper fills this gap.

@@

\hypertarget{limitations-and-practicality}{%
\section{Limitations and
practicality}\label{limitations-and-practicality}}

\begin{itemize}
\item
  explain how the lineup protocol can be cheaply done with the R package
  ``nullabor''. everyone can do it easily. The cost of using lineup in
  stat education under classroom setting is cheap (adam loy teaching
  paper)
\item
  Formal tests. reliable results. crowd sourcing is not expensive
  compared to the cost other devices in the lab and is not entirely
  unreasonable. Only if there is grey area (you can not confidently
  detect the data).
\item
  The possiblity of computer vision in the future.
\item
  repeat the scope of the study
\item
  list other things we could have tested
\item
  explain why the result should apply to other residual diagnostics
  settings
\item
  finding interesting feature that was not intended to test for
\end{itemize}
